---
title: "Organizing projected ROMS outputs for SDMs"
author: "Jameal Samhouri"
date: "5/24/2021"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(tidync)
library(here)
#library(zoo)
library(lubridate)
library(here)
library(ncdf4)
library(raster)
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose

The purpose of this script is to import and organize ROMS environmental outputs from downscaled projections, and convert them into rasters.

As of May 11, 2021, there are 6 ROMS projection files, reporting either bottom temperature or bottom oxygen associated with either the Hadley, GFDL, or IPSL GCM projections.

# Organize Projected Data into Rasters

The projected data corresponding to each of the 3 GCMs (Hadley, GFDL, IPSL) contain either bottom temperature or oxygen, for the entire ROMS grid, and for every day from 1980-2100. Note: 1980-2010 are not observed values. 1980-present should not be compared to observations since the interannual variability will not match up (by design). 

Files provided by Mike Jacox. Use beyond the Packard groundfish climate change project requires permission by Mike Jacox (michael.jacox@noaa.gov)

I adapted the code below from Steph Brodie's original work. https://github.com/stephbrodie1/WRAP_Location_CaseStudy/blob/master/Create_ROMS_Rasters.R

The first step is to convert ROMS GCM netcdf files into rasters. We have done this 2 ways: (1) using a template for the ROMS grid to correct the issue caused by the raster() function, where it does not interpret the ncdf values as midpoints of each cell even though they are midpoints, and (2) without using the template to avoid the need for interpolation of values.

NOTE: this is VERY computationally expensive and should only be run once, if at all.

```{r}
# projected netCDFs


#6 variables: lon, lat, year, month, day, covariate. the 6th variable is the covariate
#2 covariates: bottom temp, bottom oxygen
#3 global climate models: Hadley, IPSL, GFDL

n_cov_models <- 6 # 2 covariates * 3 models
cov <- 6 # position of the covariate of interest in the ncdf
gcm_options <- c("gfdl", "had", "ipsl")

#Dimensions:
#1452 years: 1980 - 2100 (121 years * 12 months)
#1452 months: 1 - 12 (121 years * 12 months)
#44195 days: 1980-2100 (121 years * 365.2479 days per year)
#181 lats: 30 - 48 degrees north @ 0.1 degree resolution
#186 lons: 115.5 - 134 degrees west @ 0.1 degree resolution

n_yr_mth_days <- 44195


#-----Load in projection output files----
#note: download files locally

files_long <- list.files(here::here('data','Projection NCDFs'),full.names = TRUE)
files_short <- list.files(here::here('data','Projection NCDFs'),full.names = FALSE)

t <- proc.time()

### NOTE ###
# break this for loop into 2 chunks so it doesn't blow up your computer. so first do for(f in c(1:3)) and then, after a restart, for(f in c(4:6))
for(f in c(4:6)){ #for (f in c(1:n_cov_models))
  print(files_long[f])
  nc <- nc_open(files_long[f])
  
  # print(nc) #run if you want more details on file
  lat <- ncvar_get(nc, 'lat'); lat <- lat[1,]
  lon <- ncvar_get(nc, 'lon'); lon <- lon[,1]
  year <- ncvar_get(nc, 'year') 
  month <- ncvar_get(nc, 'month')
  day <- ncvar_get(nc,'day')
  roms_ymd <- as.POSIXct(paste(year,month,day,sep='-'),tz='UTC')
  name <-  names(nc$var)[cov]
  
  # this step is slow. 169.131 s on Jameal's iMac
  # t2 <- proc.time()
  tmp.array <- ncvar_get(nc,name)
  # print(paste0("time to open ncdf: ", proc.time()-t2['elapsed'][[1]]))
  
  #Forcing resolution our ROMS template (a quick fix to resolve grid edge vs. mid point)
  #template <- raster(here::here('data','Bathymetry ETOPO1','template.grd')) 

  #Loop through every time step in ncdf and make a raster file.
  
  # dir.create(here::here('data','Projection Rasters','Rasters_2d_daily'),showWarnings = FALSE)
    dir.create(here::here('data','Projection Rasters','Rasters_2d_daily_notemplate'),showWarnings = FALSE)
  
  for (i in 1:n_yr_mth_days){ #1:3){
    # t3 <- proc.time()
    #create nested folders to save files
    #First create GCM folder
    gcm_name <- unlist(strsplit(files_short[f],"_"))
    index <- which(gcm_name %in% gcm_options)
    gcm_folder <- gcm_name[index]
    # dir.create(here::here('data','Projection Rasters','Rasters_2d_daily',gcm_folder), showWarnings = FALSE)
    dir.create(here::here('data','Projection Rasters','Rasters_2d_daily_notemplate',gcm_folder), showWarnings = FALSE)
    
    var <- unlist(strsplit(files_short[f],"_"))[1:2]
    # folder <- here::here('data','Projection Rasters','Rasters_2d_daily',gcm_folder,paste(var[1],var[2],sep="_"))
    folder <- here::here('data','Projection Rasters','Rasters_2d_daily_notemplate',gcm_folder,paste(var[1],var[2],sep="_"))
    dir.create(folder, showWarnings = FALSE)
    
    # now make the raster
    r <- raster(t(tmp.array[,,i]), # transposing puts lat in the rows, long in the columns
                xmn=min(lon), xmx=max(lon),
                ymn=min(lat), ymx=max(lat), 
                crs=CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))
    r <- flip(r,2) # puts lower latitudes on the bottom
    # r <- raster::resample(r, template, method="bilinear")  # method="ngb" does not change raw values from ncdf. commented out because we lose a dimension from this step. could choose a few cells and compare values 
    # plot(r)
    
    #create name of raster file
    #month_num <- format(as.Date(paste("1666",month[i],"06",sep="-"),"%Y-%m-%d"), "%m") # 1666 and 6 are arbitrary choices, goal is just to extract a 2 digit month
    fname <- paste0(var[1],"_",var[2],"_",gcm_folder,"_date_",roms_ymd[i],".grd")
    
    #save raster
    writeRaster(r, paste0(folder,"/", fname), overwrite=TRUE)
    rm(r)
    print(paste0("finished: ", fname))
  }
  nc_close(nc)
  # print(paste0("time to complete raster generation for 1 model-covariate: ", proc.time()-t3))
}
proc.time()-t

# looping through 3 times took ~1.1hr, but 2nd loop through 3 times took ~3hrs
#    user   system  elapsed 
# 5956.712  617.898 4011.009

# GRAVEYARD ###
# all code in this chunk below here is legacy of attempts by jameal to unpack the ncdf's

# 
# ## bottom temperature
# t_gfdl_file <- tidync(here::here('data','bt_daily_roms_gfdl_1980_2100.nc')) #tidync("/Users/jameal.samhouri/Documents/Future Seas/bt_daily_roms_gfdl_1980_2100.nc")
# t_hadl_file <- tidync(here::here('data','bt_daily_roms_had_1980_2100.nc'))
# t_ipsl_file <- tidync(here::here('data','bt_daily_roms_ipsl_1980_2100.nc'))
# 
# ## bottom oxygen
# o_gfdl_file <- tidync(here::here('data','oxygen_bottom_daily_roms_gfdl_1980_2100.nc'))
# o_hadl_file <- tidync(here::here('data','oxygen_bottom_daily_roms_had_1980_2100.nc'))
# o_ipsl_file <- tidync(here::here('data','oxygen_bottom_daily_roms_ipsl_1980_2100.nc'))

#glimpse(t_gfdl_file)

# # use this to determine varname. hoping we can varname="bt"
# tidync(t_gfdl_file)
# print(t_gfdl_file) #https://ropensci.org/blog/2019/11/05/tidync/ # we wantbt from 1st grid, lon and lat from 2nd grid, year/month/day from 3rd grid
# 
# nc.data <- nc_open(here::here('data','bt_daily_roms_gfdl_1980_2100.nc'), write=FALSE)
# lat <- ncvar_get(nc.data,'lat'); lat <- lat[1,]
# lon <- ncvar_get(nc.data,'lon'); lon <- lon[,1]
# yr <- ncvar_get(nc.data,'year')
# mth <- ncvar_get(nc.data,'month')
# day <- ncvar_get(nc.data,'day')
# tim <- as.POSIXct(paste(yr,mth,day,sep='-'),tz='UTC')
# name <-  names(nc.data$var)[6]
# 
# t <- proc.time()
# tmp.array <- ncvar_get(nc.data,name)
# proc.time()-t
# 
# 
# # extract netcdf files as expanded tables. takes 15-20min on jameal's imac
# t <- proc.time()
# 
# # this approach gave a tibble with columns bt, longitude, latitude, time. lon/lat were index values, time was a year
# t_gfdl <- t_gfdl_file %>% 
#   activate("D0,D1,D2") %>%
#   hyper_tibble()
# 
# proc.time()-t
# 
# glimpse(t_gfdl)
# 
# t_gfdl2 <- t_gfdl_file %>% 
#   #activate("D0,D1") %>%
#   hyper_tibble()
# 
# # # make it smaller
# # t_gfdl_slice <- t_gfdl %>% slice(1:100000)
# # rm(t_gfdl, t_gfdl_file)
# # write_rds(t_gfdl_slice, here::here('data','t_gfdl_slice.rds'))
# 
# t_gfdl_slice <- read_rds(here::here('data','t_gfdl_slice.rds'))
# glimpse(t_gfdl_slice)


```

# Derive Environmental Variables - Projections

We want to derive, from the ROMS projection outputs, some summarized environmental variables. We write some loops that can calculate, for each grid cell in the ROMS domain, a mean/97.5%/2.5% value for an environmental variable, at that location, between the files_survey_months_pattern, where files_survey_months_pattern can be anything (but months in the range May-October make the most sense because that is when the trawl survey occurs). 

Since time is just a year right now, we take the mean for each year. 

## Make annual rasters

```{r}
 # start at 0629
#----Create May-Oct Average Conditions----
#For each year and variable, load in daily data from May-Oct and average. Output to a new folder. 

years <- seq(1980,2100,1)
nyears <- length(years)
gcm_options <- c("gfdl", "had", "ipsl")
variables <- c("oxygen_bottom","bt_daily")
files_survey_months_pattern <- c("-05-","-06-","-07-","-08-","-09-","-10-") # trawl survey months are May-Oct 


for (k in 1:length(variables)){
  
  print(variables[k])
  
  for (i in 1:length(gcm_options)){
    
    print(gcm_options[i])
          
    # files <- list.files(paste0(here::here('data','Projection Rasters','Rasters_2d_daily',gcm_options[i],variables[k])), pattern=".grd" , full.names = TRUE)
    files <- list.files(paste0(here::here('data','Projection Rasters','Rasters_2d_daily_notemplate',gcm_options[i],variables[k])), pattern=".grd" , full.names = TRUE)
    
    files_survey_months <- files[Reduce("|", lapply(files_survey_months_pattern, function(x) grepl(x, files)))] #https://stackoverflow.com/questions/50681248/filter-a-list-of-files-by-date-string
    
    ### NOTE: COME BACK TO CHECK THAT THE LINE ABOVE PULLS ONLY THE CORRECT FILES
    
    start_indx <- 1979
    
    for(j in 1: nyears){
      
      # NEXT STEP IS TO SUBSET TO FILES FOR EACH YEAR ONLY
      
      files_survey_months_tmp <- files_survey_months[Reduce("|", lapply(as.character(j+start_indx), function(x) grepl(x, files_survey_months)))]
      
      # LOAD RASTERS AS A STACK 
      survey_months_rasters <- stack(files_survey_months_tmp)
      
      survey_r <- mean(survey_months_rasters)
      
      # dir.create(here::here('data','Projection Rasters','Rasters_2d_Survey_Months'), showWarnings = FALSE)
      dir.create(here::here('data','Projection Rasters','Rasters_2d_Survey_Months_notemplate'), showWarnings = FALSE)
      # writeRaster(survey_r,here::here('data','Projection Rasters','Rasters_2d_Survey_Months',paste0(gcm_options[i],'_',variables[k],'_SurveyMonths_',years[j],'.grd')), overwrite=TRUE)
      writeRaster(survey_r,here::here('data','Projection Rasters','Rasters_2d_Survey_Months_notemplate',paste0(gcm_options[i],'_',variables[k],'_SurveyMonths_',years[j],'.grd')), overwrite=TRUE)
    }
  }
}

# t_gfdl_slice_summary <- t_gfdl_slice %>%
#   mutate(
#     grid_id = paste0(latitude,"_",longitude)
#   ) %>%
#   group_by(grid_id, time) %>%
#   summarise(
#     bt_mean = mean(bt),
#     bt_upper = quantile(bt, probs=(0.975)),
#     bt_lower = quantile(bt, probs=(0.025))
#   )
# glimpse(t_gfdl_slice_summary)

```

