---
title: "Organizing ROMS hindcast data for SDM fitting"
author: "Owen Liu"
date: "5/11/2021"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(tidync)
library(here)
library(zoo)
library(lubridate)
library(here)
# library(ncdf4)
# library(raster)
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose

The purpose of this script is to import and organize ROMS environmental data, and join it to west coast trawl survey data through customizable scripts for choosing species and derived environmental variables of interest (e.g., 1 month-lagged bottom temperature).

As of May 11, 2021, there are 6 ROMS projection files, reporting either bottom temperature or bottom oxygen associated with either the Hadley, GFDL, or IPSL GCM projections.

# Organize Hindcast Data

The hindcast data includes a timeseries of bottom temperature and bottom oxygen matched to trawl survey locations

```{r}
# trawl survey locations and times
trawl <- read_csv(here::here('data','trawl_survey_locations.csv'),col_types = 'cddddd') %>% 
  mutate(date=as_date(date))

# hindcast netCDF
hindcast_file <- tidync(here::here('data','trawl_roms_matchup_timeseries.nc'))

```

```{r}
# matched ROMS hindcast data
hindcast <- hindcast_file %>% hyper_tibble()

glimpse(hindcast)

# and the trawl stations
trawlID <- hindcast_file %>% activate("D0") %>% hyper_tibble()
glimpse(trawlID)

# join trawl info to hindcast data
hindcast <- hindcast %>% left_join(trawlID, by="station")
glimpse(hindcast)
```

# Derive Environmental Variables - Hindcast

We want to derive, from the trawl-matched ROMS outputs, some summarized environmental variables. We start by writing a function that can calculate, for each trawl location and time, a mean value for an environmental variable, at that location, over the past x days, where x can be anything. For the edge case at the beginning of the timeseries, to avoid losing data we will use the 'mean of the past x days or since the beginning of the timeseries.

Since time is an integer day, it make it easier for us to calculate lags. For this function, we assume the dataframe has the exact variable names as the `hindcast` dataframe above.

For calculating lagged means, we use the `zoo::rollapply` function.
```{r}
# using `zoo::rollapply`
append_lag <- function(df,variable=temp_roms,lagdays=30){
  df_out <- df %>% 
    group_by(station) %>% # group the observations by trawl survey station
    arrange(time) %>% 
    # calculate the lag
    mutate("mean_{{variable}}_{{lagdays}}":= rollapply({{variable}}, # variable to lag
                                                       FUN=mean, # function to apply
                                                       align='right', # tells the function to lag, not lead
                                                       width=lagdays, # how big of a window
                                                       fill=NA, # value to fill when an observation doesn't exist
                                                       partial=1, # minimum window size for partial computations
                                                       na.rm=T)) %>% 
    ungroup()
  
  df_out
}
```

A quick test
```{r}
testdf <- hindcast %>% slice(1:100000)
test <- append_lag(testdf)

glimpse(test)
```

Apply to temperature and oxygen for the entire dataset

```{r}
t <- proc.time()
hindcast_lagged_temp_oxy <- hindcast %>% 
  append_lag(variable=temp_roms,lagdays=30) %>% 
  append_lag(variable=oxygen_roms,lagdays=30) %>% 
  ungroup()
proc.time()-t

# took about 25m to calculate both lags
glimpse(hindcast_lagged_temp_oxy)
```

How does concurrent temp/oxygen compare to lagged mean temp/oxy?

```{r}
# take a random subsample so it doesn't take forever to plot
samp <- hindcast_lagged_temp_oxy %>% 
  ungroup() %>% 
  slice_sample(n=10000)
samp %>% 
  ggplot(aes(temp_roms,mean_temp_roms_30))+
  geom_point()+geom_smooth(method='lm')+
  labs(x="Concurrent Temperature",y="30 Day Lagged Mean Temperature")
```



