---
title: "TidyNC Projected ROMS"
author: "Owen Liu"
date: "6/2/2021"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(tidync)
library(here)
library(lubridate)
library(here)
# library(ncdf4)
# library(raster)
knitr::opts_chunk$set(echo = TRUE)
 options(dplyr.summarise.inform = FALSE)
```

# Purpose

Try a different way of extracting and organizing projected ROMS data for use in species distribution models. The main idea is to test whether using `tidync` can make the process smoother and faster (i.e., fewer translation steps between ncdf, raster, and tabular data formats)

# Import Data

We'll just test with one of the projection files.

```{r}
bt_gfdl_fl <- here('data','Projection NCDFs','bt_daily_roms_gfdl_1980_2100.nc')
bt_gfdl <- tidync(bt_gfdl_fl)
print(bt_gfdl)
```

```{r}
# look at the grids
hyper_grids(bt_gfdl)
hyper_dims(bt_gfdl)
hyper_vars(bt_gfdl)
```


The netCDFs have longitude, latitude, and time (year, month, day, and `roms_time`, which is seconds since 1900-01-01). Goal- pull each trawl survey season (May-October) and average at each grid cell.

# Create a Time Reference

Time is encoded a bit weirdly in the ncdf, so we need to figure out a time "index" to assist with referencing between year/month/day and the actual time variable.

```{r}
# time index
time_reference <- bt_gfdl %>% activate("D2") %>% 
  hyper_tibble() %>% 
  # add an index
  mutate(idx=row_number())
glimpse(time_reference)
```

We'll use this time reference to help in filtering the dataset to extract each year's data. Here's an example:

```{r}
# find the correct "time index" for 1980, all dates between May and October inclusive
x<- time_reference %>% filter(year==1980,month>=5,month<=10) %>% 
  # extract the appropriate time index
  pluck('idx')

# try taking a slice of data for this time index
y <- bt_gfdl %>% hyper_filter(time=index%in%x) %>% hyper_tibble()
# seems to work

# now summarize means by grid cell
z <- y %>% 
  group_by(longitude,latitude) %>% 
  summarise(mean_bt=mean(bt)) %>% 
  mutate(year=1980) %>% 
  ungroup()
# seems to work
```

# Annual Summary Function

Okay this general workflow seems to work. Let's write a function and make a loop to apply this to all years in the data.

```{r}

ncdf_file <- here('data','Projection NCDFs','oxygen_bottom_daily_roms_gfdl_1980_2100.nc')
yr <- 2000

pull_season_df <- function(ncdf_file,yr){
  # load the file reference
  nc <- tidync(ncdf_file)
  
  # get the input variable name
  v <- hyper_vars(nc) %>% pluck("name")
  # output variable name
  out <- paste0("mean_",v)
  
  # find the appropriate time reference (may to october of the given year)
  tm <- time_reference %>% 
    filter(year==yr,month>=5,month<=10) %>% 
    pluck('idx')
  
  # extract and summarize the data across the season for each lat/lon
  dat <- nc %>% 
    hyper_filter(time=index%in%tm) %>%
    hyper_tibble() %>% 
    group_by(longitude,latitude) %>% 
    summarise(!!out:=mean(!!sym(v))) %>% 
    mutate(year=yr) %>% 
    ungroup()
  
  print(paste("Extracted",out,"for",yr))
  # return the summarized dataframe
  return(dat)
}
```

Test the function for an example year

```{r}
bt_gfdl_2035 <- pull_season_df(bt_gfdl_fl,yr=2035)
glimpse(bt_gfdl_2035)
```

And for oxygen

```{r}
oxygen_gfdl_2035 <- pull_season_df(here('data','Projection NCDFs','oxygen_bottom_daily_roms_gfdl_1980_2100.nc'),yr=2035)
glimpse(oxygen_gfdl_2035)
```


# Apply Summary Function

Now, we can apply the above function to all of the data.

```{r}
yrs_to_extract <- 1980:2100
```

```{r,message=F}

t <- proc.time()

# for GFDL bottom temperature
bt_gfdl_summ <- purrr::map_df(yrs_to_extract,pull_season_df,ncdf_file=bt_gfdl_fl)

# for IPSL bottom temperature
bt_ipsl_fl <- here('data','Projection NCDFs','bt_daily_roms_ipsl_1980_2100.nc')
bt_ipsl_summ <- purrr::map_df(yrs_to_extract,pull_season_df,ncdf_file=bt_ipsl_fl)

# for Hadley bottom temperature
bt_had_fl <- here('data','Projection NCDFs','bt_daily_roms_had_1980_2100.nc')
bt_had_summ <- purrr::map_df(yrs_to_extract,pull_season_df,ncdf_file=bt_had_fl)

# for GFDL bottom oxygen
oxygen_bottom_gfdl_fl <- here('data','Projection NCDFs','oxygen_bottom_daily_roms_gfdl_1980_2100.nc')
oxygen_bottom_gfdl_summ <- purrr::map_df(yrs_to_extract,pull_season_df,ncdf_file=oxygen_bottom_gfdl_fl)

# for IPSL bottom oxygen
oxygen_bottom_ipsl_fl <- here('data','Projection NCDFs','oxygen_bottom_daily_roms_ipsl_1980_2100.nc')
oxygen_bottom_ipsl_summ <- purrr::map_df(yrs_to_extract,pull_season_df,ncdf_file=oxygen_bottom_ipsl_fl)

# for Hadley bottom oxygen
oxygen_bottom_had_fl <- here('data','Projection NCDFs','oxygen_bottom_daily_roms_had_1980_2100.nc')
oxygen_bottom_had_summ <- purrr::map_df(yrs_to_extract,pull_season_df,ncdf_file=oxygen_bottom_had_fl)

proc.time()-t
```

# Join summary from individual models into a single data frame

```{r}

df_summ <- bt_gfdl_summ %>%
  rename(mean_bt_gfdl = mean_bt) %>%
  left_join(
    bt_ipsl_summ %>% rename( mean_bt_ipsl = mean_bt)
  ) %>%
  left_join(
    bt_had_summ %>% rename( mean_bt_hadl = mean_bt)
  ) %>%
  left_join(
    oxygen_bottom_gfdl_summ %>% rename( mean_oxy_bottom_gfdl = mean_oxy_bottom)
  ) %>%
  left_join(
    oxygen_bottom_ipsl_summ %>% rename( mean_oxy_bottom_ipsl = mean_oxy_bottom)
  ) %>%
  left_join(
    oxygen_bottom_had_summ %>% rename( mean_oxy_bottom_hadl = mean_oxy_bottom)
  ) 
glimpse(df_summ)

```

# Replace index values for lat and long with midpoint coordinates

```{r}

ll_ref <- bt_gfdl %>% activate("D0,D1") %>% hyper_tibble()

df_summ <- df_summ %>%
  left_join(ll_ref) %>%
  select(
    year, lat, lon, everything()
  ) %>%
  select(
    -longitude, -latitude
  )
glimpse(df_summ)

```

# Write out the summary data frame

```{r}

write_rds(df_summ,here('data','mean_bt_ox_allmodels_1980_2100.rds'))

```