---
title: "TidyNC Projected ROMS"
author: "Owen Liu"
date: "6/2/2021"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(tidync)
library(here)
library(lubridate)
library(here)
library(zoo)
# library(ncdf4)
# library(raster)
knitr::opts_chunk$set(echo = TRUE)
 options(dplyr.summarise.inform = FALSE)
```

# Purpose

Try a different way of extracting and organizing projected ROMS data for use in species distribution models. The main idea is to test whether using `tidync` can make the process smoother and faster (i.e., fewer translation steps between ncdf, raster, and tabular data formats)

# Import Data

We'll just test with one of the projection files.

```{r}
bt_gfdl_fl <- here('data','Projection NCDFs','bt_daily_roms_gfdl_1980_2100.nc')
bt_gfdl <- tidync(bt_gfdl_fl)
print(bt_gfdl)
```

```{r}
# look at the grids
hyper_grids(bt_gfdl)
hyper_dims(bt_gfdl)
hyper_vars(bt_gfdl)
```


The netCDFs have longitude, latitude, and time (year, month, day, and `roms_time`, which is seconds since 1900-01-01). Goal- pull each trawl survey season (May-October) and average at each grid cell.

# Create a Time Reference

Time is encoded a bit weirdly in the ncdf, so we need to figure out a time "index" to assist with referencing between year/month/day and the actual time variable.

```{r}
# time index
time_reference <- bt_gfdl %>% activate("D2") %>% 
  hyper_tibble() %>% 
  # add an index
  mutate(idx=row_number())
glimpse(time_reference)
```

We'll use this time reference to help in filtering the dataset to extract each year's data. Here's an example:

```{r}
# find the correct "time index" for 1980, all dates between May and October inclusive
x<- time_reference %>% filter(year==1980,month>=5,month<=10) %>% 
  # extract the appropriate time index
  pluck('idx')

# try taking a slice of data for this time index
y <- bt_gfdl %>% hyper_filter(time=index%in%x) %>% hyper_tibble()
# seems to work

# now summarize means by grid cell
z <- y %>% 
  group_by(longitude,latitude) %>% 
  summarise(mean_bt=mean(bt)) %>% 
  mutate(year=1980) %>% 
  ungroup()
# seems to work

# figure out the index value for July 31, the midpoint of the survey months = 92
# take a slice of data for this time index
midsurvey_index <- min(x) + 91
y2 <- bt_gfdl %>% hyper_filter(time=index%in%midsurvey_index) %>% hyper_tibble()

# calculate the 30d avg bt leading up to july 31. JS adapted this from the append_lag() function Owen wrote for prepare_hindcast_ROMS.Rmd
lagdays=30
z2 <- y2 %>% 
  group_by(longitude,latitude) %>% 
  summarise(
    mean_bt_30d = rollapply(bt, # variable to lag
                                                       FUN=mean, # function to apply
                                                       align='right', # tells the function to lag, not lead
                                                       width=lagdays, # how big of a window
                                                       fill=NA, # value to fill when an observation doesn't exist
                                                       partial=1, # minimum window size for partial computations
                                                       na.rm=T)) %>% 
  mutate(year=1980) %>% 
  ungroup()
# seems to work

```

# Annual Summary Function

Okay this general workflow seems to work. Let's write a function and make a loop to apply this to all years in the data. The resulting tibble will include the average value of the environmental variable each year, along with the average value of that environmental variable over the 30d leading up to July 31 (the midpoint of the trawl survey season) each year.

```{r}

# ncdf_file <- here('data','Projection NCDFs','oxygen_bottom_daily_roms_gfdl_1980_2100.nc')
# yr <- 2000

pull_season_df <- function(ncdf_file, yr, lagdays=30, lagdaysintosurvey=91){
  # load the file reference
  nc <- tidync(ncdf_file)
  
  # get the input variable name
  v <- hyper_vars(nc) %>% pluck("name")
  # output variable name
  out <- paste0("mean_",v)
  out2 <- paste0("mean_",v,"_30d")
  
  # find the appropriate time reference (may to october of the given year)
  tm <- time_reference %>% 
    filter(year==yr,month>=5,month<=10) %>% 
    pluck('idx')
  
  # figure out the index value for July 31, the midpoint of the survey months = 92
  # take a slice of data for this time index
  midsurvey_index <- min(tm) + lagdaysintosurvey
  
  # extract and summarize the data across the season for each lat/lon
  dat <- nc %>% 
    hyper_filter(time=index%in%tm) %>%
    hyper_tibble() %>% 
    group_by(longitude,latitude) %>% 
    summarise(
      !!out:=mean(!!sym(v))
    ) %>% 
    mutate(year=yr) %>% 
    ungroup()
  
  dat1 <- nc %>% 
    hyper_filter(time=index%in%midsurvey_index) %>%
    hyper_tibble() %>% 
    group_by(longitude,latitude) %>% 
    summarise(
      !!out2:= rollapply(!!sym(v), # variable to lag
                                                       FUN=mean, # function to apply
                                                       align='right', # tells the function to lag, not lead
                                                       width=lagdays, # how big of a window
                                                       fill=NA, # value to fill when an observation doesn't exist
                                                       partial=1, # minimum window size for partial computations
                                                       na.rm=T)
      ) %>% 
    mutate(year=yr) %>% 
    ungroup()
  
  dat2 <- dat %>%
    left_join(dat1)
  
  print(paste("Extracted",out,"for",yr))
  # return the summarized dataframe
  return(dat2)
}
```

Test the function for an example year

```{r}
bt_gfdl_2035 <- pull_season_df(bt_gfdl_fl,yr=2035)
glimpse(bt_gfdl_2035)
```

And for oxygen

```{r}
oxygen_gfdl_2035 <- pull_season_df(here('data','Projection NCDFs','oxygen_bottom_daily_roms_gfdl_1980_2100.nc'),yr=2035)
glimpse(oxygen_gfdl_2035)
```


# Apply Summary Function

Now, we can apply the above function to all of the data.

```{r}
yrs_to_extract <- 1980:2100
```

```{r,message=F}

t <- proc.time()

# for GFDL bottom temperature
bt_gfdl_summ <- purrr::map_df(yrs_to_extract,pull_season_df,ncdf_file=bt_gfdl_fl)

# for IPSL bottom temperature
bt_ipsl_fl <- here('data','Projection NCDFs','bt_daily_roms_ipsl_1980_2100.nc')
bt_ipsl_summ <- purrr::map_df(yrs_to_extract,pull_season_df,ncdf_file=bt_ipsl_fl)

# for Hadley bottom temperature
bt_had_fl <- here('data','Projection NCDFs','bt_daily_roms_had_1980_2100.nc')
bt_had_summ <- purrr::map_df(yrs_to_extract,pull_season_df,ncdf_file=bt_had_fl)

# for GFDL bottom oxygen
oxygen_bottom_gfdl_fl <- here('data','Projection NCDFs','oxygen_bottom_daily_roms_gfdl_1980_2100.nc')
oxygen_bottom_gfdl_summ <- purrr::map_df(yrs_to_extract,pull_season_df,ncdf_file=oxygen_bottom_gfdl_fl)

# for IPSL bottom oxygen
oxygen_bottom_ipsl_fl <- here('data','Projection NCDFs','oxygen_bottom_daily_roms_ipsl_1980_2100.nc')
oxygen_bottom_ipsl_summ <- purrr::map_df(yrs_to_extract,pull_season_df,ncdf_file=oxygen_bottom_ipsl_fl)

# for Hadley bottom oxygen
oxygen_bottom_had_fl <- here('data','Projection NCDFs','oxygen_bottom_daily_roms_had_1980_2100.nc')
oxygen_bottom_had_summ <- purrr::map_df(yrs_to_extract,pull_season_df,ncdf_file=oxygen_bottom_had_fl)

proc.time()-t
```

# Join summary from individual models into a single data frame

```{r}

df_summ <- bt_gfdl_summ %>%
  rename(mean_bt_gfdl = mean_bt,
         mean_bt_30d_gfdl = mean_bt_30d) %>%
  left_join(
    bt_ipsl_summ %>% 
      rename( mean_bt_ipsl = mean_bt,
              mean_bt_30d_ipsl = mean_bt_30d)
  ) %>%
  left_join(
    bt_had_summ %>% 
      rename( mean_bt_hadl = mean_bt,
              mean_bt_30d_hadl = mean_bt_30d)
  ) %>%
  left_join(
    oxygen_bottom_gfdl_summ %>% 
      rename( mean_oxy_bottom_gfdl = mean_oxy_bottom,
              mean_oxy_bottom_30d_gfdl = mean_oxy_bottom_30d)
  ) %>%
  left_join(
    oxygen_bottom_ipsl_summ %>% 
      rename( mean_oxy_bottom_ipsl = mean_oxy_bottom,
              mean_oxy_bottom_30d_ipsl = mean_oxy_bottom_30d)
  ) %>%
  left_join(
    oxygen_bottom_had_summ %>% 
      rename( mean_oxy_bottom_hadl = mean_oxy_bottom,
              mean_oxy_bottom_30d_hadl = mean_oxy_bottom_30d)
  ) 
glimpse(df_summ)

```

# Replace index values for lat and long with midpoint coordinates

```{r}

ll_ref <- bt_gfdl %>% activate("D0,D1") %>% hyper_tibble()

df_summ <- df_summ %>%
  left_join(ll_ref) %>%
  select(
    year, lat, lon, everything()
  ) %>%
  select(
    -longitude, -latitude
  )
glimpse(df_summ)

```

# Write out the summary data frame

```{r}

# write_rds(df_summ,here('data','mean_bt_ox_allmodels_1980_2100.rds'))
write_rds(df_summ,here('data','mean_30dmean_bt_ox_allmodels_1980_2100.rds'))

```