---
title: "Visualize Annual Environmental Data"
author: "Owen Liu, Jameal Samhouri"
date: "5/25/2021"
output: html_document
---


```{r setup, include=FALSE}
library(tidyverse)
library(tidync)
library(here)
library(zoo)
library(lubridate)
library(here)
# library(ncdf4)
library(raster)
library(tabularaster)
library(viridis)
library(ggsci)
library(cmocean)
library(maps)
library(ggpubr)
knitr::opts_chunk$set(echo = TRUE)
options(dplyr.summarise.inform=F)
```

# Purpose

Take the rasterized annual outputs from projected ROMS and visualize them with maps and timeseries. The environmental data have already been extracted and summarized for temperature and oxygen, such that the value for each cell has either mean bottom temperature or mean bottom oxygen, across the season from May to October.

There are 6 projection types total, corresponding to three GCMs (IPSL, Hadley, and GFDL) and two environmental variables (temperature and oxygen)

# Import Data

Import the raster data into stacks

```{r}
# organize the file names we want to import
# x <- list.files(here('data','Projection Rasters','Rasters_2d_Survey_Months'),full.names=T,pattern = ".grd")
# gfdl_bt_fl <- x[grepl('gfdl_bt',x)]
# gfdl_o_fl <- x[grepl('gfdl_oxygen',x)]
# had_bt_fl <- x[grepl('had_bt',x)]
# had_o_fl <- x[grepl('had_oxygen',x)]
# ipsl_bt_fl <- x[grepl('ipsl_bt',x)]
# ipsl_o_fl <- x[grepl('ipsl_oxygen',x)]
```

Import the rds file containing the direct extraction and summary values from the ncdf files. Grab the bathymetry layer too and merge it in, then drop all cells with depths >1280m.

```{r}

df_summ <- read_rds(here('data','Projection Summaries','mean_bt_ox_allmodels_1980_2100.rds'))

# df_summ %>%
#   group_by(year, lat) %>%
#   summarise(
#     mean_bt_gfdl = mean(mean_bt_gfdl)
#   ) %>%
#   filter( lat == 30.0 )

bathy_fl <-  here('data','bottom_layer_depth.nc')
bathy <- tidync(bathy_fl)
print(bathy)
ncmeta::nc_grids(bathy)
ncmeta::nc_vars(bathy_fl)
ncmeta::nc_atts(bathy_fl, "bld") %>% tidyr::unnest(cols = c(value)) # tells us bld data are in m

bathy_df <- bathy %>% activate("D0,D1") %>% hyper_tibble() #bld: bottom layer depth, units for bld are meters, and larger values mean deeper

df_summ_gfish <- df_summ %>%
  left_join(bathy_df) %>%
  dplyr::select(-longitude, -latitude) %>%
  filter(
    bld <= 1280
  )
glimpse(df_summ_gfish)  

```

# Visualize

We'll try a few different visualizations, writing functions for each that:

* import a raster stack using a list of file names (created above)
* summarize the data somehow
* output a plot or return a dataframe

## Mean Conditions

```{r}
# mean_raster <- function(file_list){
#   # import files into a raster stack
#   s <- stack(file_list)
#   # take the mean
#   m <- mean(s)
#   # label for plot
#   pt <- case_when(
#     grepl('ipsl_oxygen',file_list[1]) ~ "IPSL Oxygen",
#     grepl('ipsl_bt',file_list[1]) ~ "IPSL Temperature",
#     grepl('gfdl_oxygen',file_list[1]) ~ "GFDL Oxygen",
#     grepl('gfdl_bt',file_list[1]) ~ "GFDL Temperature",
#     grepl('had_oxygen',file_list[1]) ~ "HAD Oxygen",
#     grepl('had_bt',file_list[1]) ~ "HAD Temperature",
#   )
#   # plot
#   plot(m,main=pt,col=cividis(256))
# }
```
Try it out

```{r}
# purrr::map(list(gfdl_bt_fl,had_bt_fl,ipsl_bt_fl,gfdl_o_fl,had_o_fl,ipsl_o_fl), mean_raster)
```

## Time Series

First, a latitude-binned time series of mean values from the raster stack

```{r}
# plot_ts <- function(file_list,return_plot=T){
#   # import files into a raster stack
#   s <- stack(file_list)
#   yrs <- 1980:2100
#   # convert to tibble
#   sdf <- s %>% as_tibble(xy=T) %>% 
#     #split into bins
#     mutate(bin=findInterval(y,vec=seq(30,50,by=5)),
#            year=yrs[dimindex]) %>% 
#     # summarize by latitude bin
#     group_by(year,bin) %>% 
#     summarise(m=mean(cellvalue,na.rm=T)) %>% 
#     ungroup()
#   if(return_plot){
#     # create labels for latitude bins
#     sdf <- sdf %>% 
#       mutate(latbin=case_when(
#         bin==1 ~ "30-35",
#         bin==2 ~ "35-40",
#         bin==3 ~ "40-45",
#         bin==4 ~ ">45"
#       ))
#     # label for plot
#     pt <- case_when(
#       grepl('ipsl_oxygen',file_list[1]) ~ "IPSL Oxygen",
#       grepl('ipsl_bt',file_list[1]) ~ "IPSL Temperature",
#       grepl('gfdl_oxygen',file_list[1]) ~ "GFDL Oxygen",
#       grepl('gfdl_bt',file_list[1]) ~ "GFDL Temperature",
#       grepl('had_oxygen',file_list[1]) ~ "HAD Oxygen",
#       grepl('had_bt',file_list[1]) ~ "HAD Temperature",
#     )
#     lab <- ifelse(grepl('bt',file_list[1]),"Bottom Temperature","Bottom Oxygen")
#     # make plot
#     p <- sdf %>% 
#       ggplot(aes(year,m,col=fct_reorder(latbin,bin)))+
#       geom_line()+
#       scale_color_locuszoom()+
#       labs(x='Year',y=paste("Mean",lab),col="Latitude",title=pt)+
#       theme_minimal()
#     return(p)
#   } else return(sdf)
# }
```

Try it out

```{r}
# purrr::map(list(gfdl_bt_fl,had_bt_fl,ipsl_bt_fl,gfdl_o_fl,had_o_fl,ipsl_o_fl), plot_ts)
```

Second, a latitude-binned time series of mean values from the direct extraction

```{r}
# df_name <- df_summ_gfish
# var_name <- "mean_bt_gfdl"
plot_ts_direct <- function(df_name, var_name, return_plot=T){
  
  #df_name$year = as.factor(df_name$year) # tried this to fix and did not help
  
  sdf <- df_name %>% #as_tibble(xy=T) %>% 
    #split into bins
    mutate(bin=findInterval(lat,vec=seq(30,50,by=5))) %>% 
    # summarize by latitude bin
    group_by(year, bin) %>% 
    summarise( #m2 = mean(mean_bt_gfdl))
      m=mean(.data[[var_name]],na.rm=T)
      ) %>% # this is producing a value that does not change between 1980-2100
    ungroup()
  if(return_plot){
    # create labels for latitude bins
    sdf <- sdf %>% 
      mutate(latbin=case_when(
        bin==1 ~ "30-35",
        bin==2 ~ "35-40",
        bin==3 ~ "40-45",
        bin==4 ~ ">45"
      ))
    # label for plot
    pt <- case_when(
      grepl('oxy_bottom_ipsl',var_name) ~ "IPSL Oxygen",
      grepl('mean_bt_ipsl',var_name) ~ "IPSL Temperature",
      grepl('mean_oxy_bottom_gfdl',var_name) ~ "GFDL Oxygen",
      grepl('mean_bt_gfdl',var_name) ~ "GFDL Temperature",
      grepl('mean_oxy_bottom_hadl',var_name) ~ "HAD Oxygen",
      grepl('mean_bt_hadl',var_name) ~ "HAD Temperature",
    )
    lab <- ifelse(grepl('bt',var_name),"Bottom Temperature","Bottom Oxygen")
    # make plot
    p <- sdf %>% 
      ggplot(aes(year,m,col=fct_reorder(latbin,bin)))+
      geom_line()+
      scale_color_locuszoom()+
      labs(x='Year',y=paste("Mean",lab),col="Latitude",title=pt)+
      theme_minimal()
    return(p)
  } else return(sdf)
}
```

Try it out

```{r}
# i want to apply purrr::map  across a list of var_name using the function plot_ts_direct() with other arguments set to constant
model_var <- c("mean_bt_gfdl", "mean_bt_ipsl", "mean_bt_hadl", "mean_oxy_bottom_gfdl","mean_oxy_bottom_ipsl","mean_oxy_bottom_hadl")

purrr::map(model_var, plot_ts_direct, df_name = df_summ_gfish, return_plot=T) 
```

And save a tibble

```{r}
latbins_summ <- purrr::map(model_var, plot_ts_direct, df_name = df_summ_gfish, return_plot=F)
write_rds(latbins_summ, here::here('data','Projection Summaries','mean_bt_ox_allmodels_1980_2100_latbin.rds'))

```

Note this output is a list of 6 tibbles, one for each model-variable combination and in the order specified by `model_var`.


# Make some maps based on the directly extracted summary data

Start by looking at 1980-2010 vs 2070-2100. Then make a difference map
```{r}
# df_name <- df_summ_gfish
# var_name <- 'mean_bt_gfdl'

plot_map_direct <- function(df_name, var_name, return_plot=T){
  
  df_begin <- df_name %>% 
    filter(year<=2010) %>% 
    group_by(lat,lon) %>% 
    summarise(
      m = mean(.data[[var_name]],na.rm=T)
      ) %>%
    ungroup()
    
  df_end <- df_name %>% 
    filter(year>=2070) %>% 
    group_by(lat,lon) %>% 
    summarise(
      m = mean(.data[[var_name]],na.rm=T)
      ) %>%
    ungroup()
  
  df_delta <- df_begin %>%
    left_join(df_end, by = c('lat','lon')) %>%
    mutate(
      delta = m.y-m.x
    ) %>%
    ungroup()
  
  if(return_plot){
    # label for plot
    pt <- case_when(
      grepl('oxy_bottom_ipsl',var_name) ~ "IPSL Bottom Oxygen",
      grepl('mean_bt_ipsl',var_name) ~ "IPSL Bottom Temperature",
      grepl('mean_oxy_bottom_gfdl',var_name) ~ "GFDL Bottom Oxygen",
      grepl('mean_bt_gfdl',var_name) ~ "GFDL Bottom Temperature",
      grepl('mean_oxy_bottom_hadl',var_name) ~ "HAD Bottom Oxygen",
      grepl('mean_bt_hadl',var_name) ~ "HAD Bottom Temperature",
    )
    lab <- ifelse(grepl('bt',var_name),"Bottom\nTemperature","Bottom\nOxygen")
    # delta_ll <- ifelse(grepl('bt',var_name),
    #                        min(c(df_name$mean_bt_gfdl, df_name$mean_bt_ipsl, df_name$mean_bt_hadl)),
    #                        min(c(df_name$mean_oxy_gfdl, df_name$mean_oxy_ipsl, df_name$mean_oxy_hadl))
    # )
    # delta_ul <- ifelse(grepl('bt',var_name), 
    #                    max(c(df_name$mean_bt_gfdl, df_name$mean_bt_ipsl, df_name$mean_bt_hadl)), 
    #                    max(c(df_name$mean_oxy_gfdl, df_name$mean_oxy_ipsl, df_name$mean_oxy_hadl))
    # )
    
    # make plots
    p_begin <- df_begin %>% 
      ggplot(aes(x=lon, y=lat)) + 
      geom_tile(aes(fill=m)) +
      theme_classic() + labs(y="", x="", title = pt, subtitle = "1980-2010") + 
      theme(legend.position="right",
            legend.title = element_blank(),
            plot.title = element_text(size = 14),
            plot.subtitle = element_text(size = 12)) +
      theme( panel.border = element_rect(colour = "black", fill=NA, size=1)) + #makes a box
      scale_fill_gradientn(colours = cmocean("ice")(256), 
                           limits = c(min(df_delta$m.x), max(df_delta$m.y)),
                           name=lab) + # lab is not working
      annotation_map(map_data("world"), colour = "black", fill="grey50") +
      coord_quickmap(xlim=c(-126,-115.8),ylim=c(30,48)) +  #Sets aspect ratio
      scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0))
    
    p_end <- df_end %>% 
      ggplot(aes(x=lon, y=lat)) + 
      geom_tile(aes(fill=m)) +
      theme_classic() + labs(y="", x="", title = pt, subtitle = "2070-2100") + 
      theme(legend.position="right",
            legend.title = element_blank(),
            plot.title = element_text(size = 14),
            plot.subtitle = element_text(size = 12)) +
      theme( panel.border = element_rect(colour = "black", fill=NA, size=1)) + #makes a box
      scale_fill_gradientn(colours = cmocean("ice")(256), 
                           limits = c(min(df_delta$m.x), max(df_delta$m.y)),
                           name=lab) + # lab is not working
      annotation_map(map_data("world"), colour = "black", fill="grey50") +
      coord_quickmap(xlim=c(-126,-115.8),ylim=c(30,48)) +  #Sets aspect ratio
      scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0))
    
    p_delta <- df_delta %>% 
      ggplot(aes(x=lon, y=lat)) + 
      geom_tile(aes(fill=delta)) +
      theme_classic() + labs(y="", x="", title = pt, subtitle = "Difference between\n2070-2100 and 1980-2010") + 
      theme(legend.position="right",
            legend.title = element_blank(),
            plot.title = element_text(size = 14),
            plot.subtitle = element_text(size = 12)) +
      theme( panel.border = element_rect(colour = "black", fill=NA, size=1)) + #makes a box
      scale_fill_gradient2(
        name=lab,
        low = 'blue', mid = 'white', high = 'red',
        midpoint = 0,
        limts = c(min(df_delta$delta), max(df_delta$delta))
        #limits = delta_limits
      ) +
      # scale_fill_gradientn(colours = colorRampPalette(colors = c("#3498db", "white", "#e74c3c"))(256), 
      #                      limits = c(min(df_delta$delta), max(df_delta$delta)),
      #                      name=lab) + # lab is not working
      annotation_map(map_data("world"), colour = "black", fill="grey50") +
      coord_quickmap(xlim=c(-126,-115.8),ylim=c(30,48)) +  #Sets aspect ratio
      scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0))
    
    #return(list(p_begin, p_end, p_delta))
    return(ggarrange(p_begin,
          p_end,
          p_delta,
          ncol=3,
          nrow=1,
          legend="bottom",
          labels="auto",
          font.label = 14,
          vjust=8,
          hjust=0
          )
          )
  } else return(list(df_begin, df_end, df_delta))
}


# ggarrange(p_begin,
#           p_end,
#           p_delta,
#           ncol=3,
#           nrow=1,
#           legend="bottom",
#           labels="auto",
#           font.label = 14,
#           vjust=8,
#           hjust=0
#           )

# mean_bt_gfdl_end <- df_summ %>% filter(year>=2070) %>% group_by(lat,lon) %>% summarise(mean_bt_gfdl = mean(mean_bt_gfdl)) #%>%
  #rename(x=lat, y=lon) #%>% pivot_wider(names_from=lon,values_from = mean_bt_gfdl)

# map1 <- ggplot(data=mean_bt_gfdl_end, aes(x=lon, y=lat))+
#   geom_tile(aes(fill=mean_bt_gfdl))+
#   theme_classic() +  labs(y="", x="", title = "mean_bt_gfdl 2070-2100") +   theme(legend.position="right",legend.title = element_blank())+
#   theme( panel.border = element_rect(colour = "black", fill=NA, size=1)) + #makes a box
#   scale_fill_gradientn(colours = cmocean("ice")(256), name="Bottom Temperature") +
#   annotation_map(map_data("world"), colour = "black", fill="grey50")+
#   coord_quickmap(xlim=c(-126,-115.8),ylim=c(30,48)) +  #Sets aspect ratio
#   scale_x_continuous(expand = c(0, 0)) +  scale_y_continuous(expand = c(0, 0))
# map1


# inspired by: https://github.com/stephbrodie1/WRAP_Location_CaseStudy/blob/master/SummaryPlots.R

```

Try it out

```{r}
# i want to apply purrr::map  across a list of var_name using the function plot_ts_direct() with other arguments set to constant
model_var <- c("mean_bt_gfdl", "mean_bt_ipsl", "mean_bt_hadl", "mean_oxy_bottom_gfdl","mean_oxy_bottom_ipsl","mean_oxy_bottom_hadl")

purrr::map(model_var, plot_map_direct, df_name = df_summ_gfish, return_plot=T)
```

And save a tibble

```{r}
maps_df <- purrr::map(model_var, plot_map_direct, df_name = df_summ_gfish, return_plot=F)
write_rds(maps_df, here::here('data','Projection Summaries','mean_bt_ox_allmodels_formaps.rds'))

```

Note this output is a list of 6 tibbles, one for each model-variable combination and in the order specified by `model_var`.

# Hovmoller plots

Plot latitude and year with the variable represented by colors

Start by writing a function to summarise the the rolling average of temp and O2, by latutude and year

```{r}
# df_name <- df_summ_gfish
# lagyears=10
# var_name = "mean_bt_gfdl"
# quantile(df_summ_gfish$bld)

make_df_hov <- function(df_name, var_name, lagyears=10){
  df_hov <- df_name %>%
  filter( between (bld, 150, 300) ) %>% # choose only depths from 150-300m
  group_by(year,lat) %>%
  summarise(
    tenyrmean = rollapply(.data[[var_name]], # variable to lag
                                                       FUN=mean, # function to apply
                                                       align='right', # tells the function to lag, not lead
                                                       width=lagyears, # how big of a window
                                                       fill=NA, # value to fill when an observation doesn't exist
                                                       partial=1, # minimum window size for partial computations
                                                       na.rm=T)
  ) %>%
    mutate(latbin=29+findInterval(lat,vec=seq(30,50,by=1)))
  return(df_hov)
}

```

Now apply the function and save a list of tibbles

```{r}
# i want to apply purrr::map  across a list of var_name using the function plot_ts_direct() with other arguments set to constant
model_var <- c("mean_bt_gfdl", "mean_bt_ipsl", "mean_bt_hadl", "mean_oxy_bottom_gfdl","mean_oxy_bottom_ipsl","mean_oxy_bottom_hadl")

hov_df <- purrr::map(model_var, make_df_hov, df_name = df_summ_gfish, lagyears=10)
write_rds(hov_df, here::here('data','Projection Summaries','10yrmean_150to300m_bt_ox_allmodels_forhovmollers.rds'))
# write_rds(hov_df, here::here('data','Projection Summaries','10yrmean_bt_ox_allmodels_forhovmollers.rds'))

```

Now make some plots [still in progress]

```{r}

# # label for plot
#     pt <- case_when(
#       grepl('oxy_bottom_ipsl',var_name) ~ "IPSL Oxygen",
#       grepl('mean_bt_ipsl',var_name) ~ "IPSL Temperature",
#       grepl('mean_oxy_bottom_gfdl',var_name) ~ "GFDL Oxygen",
#       grepl('mean_bt_gfdl',var_name) ~ "GFDL Temperature",
#       grepl('mean_oxy_bottom_hadl',var_name) ~ "HAD Oxygen",
#       grepl('mean_bt_hadl',var_name) ~ "HAD Temperature",
#     )
#     lab <- ifelse(grepl('bt',var_name),"Bottom Temperature","Bottom Oxygen")


p_hov <- hov_df[[3]] %>% 
      ggplot(aes(x=year, y=latbin)) + 
      geom_tile(aes(fill=tenyrmean)) +
      theme_classic() + labs(x="Year", y="Latitude", title = pt) + 
      theme(legend.position="right",
            legend.title = element_blank(),
            plot.title = element_text(size = 14),
            plot.subtitle = element_text(size = 12)) +
      theme( panel.border = element_rect(colour = "black", fill=NA, size=1)) + #makes a box
      scale_fill_gradient2(
        name=lab,
        low = 'blue', mid = 'white', high = 'red',
        midpoint = 0,
        limits = c(min(hov_df[[3]]$tenyrmean), max(hov_df[[3]]$tenyrmean))
      )
```